---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


```{r}
# required packages
#install.packages("devtools")
#install.packages("Rtools") # Windows only?
# clean up before reloading a new version
#detach("package:chloe")
#.rs.restartR()

#remotes::install_github("pauI/chloeR") # install package
#library("chloe") # load package
#get.java() # a Java Virtual Machine is needed, to retrieve the default path

#set.java("path/to/java/java.exe") # to declare a specific JVM
```

```{r message=FALSE}
library(chloe)
library(dplyr)
library(here)
library(terra)
library(tmap)
library(RColorBrewer)
library(sf)
```

Notes: feedback for package developers

- I had to run 'eco.landscape' separately for each scale because it produced only one tif file per #classes.
- The output .tif files had correct coordinates but incorrect crs information! A French crs was assigned. I had to import the files, change their crs information, and save them under a different name.
- Cells with NA in the input map are included in the clustering, this should probably be changed. I worked around this by masking the output maps. However, this creates some artefacts where some internal areas are classified with the NA's (unless these were NA?).



# 1. Apply `eco.landscape` to lower watershed ELC map

## a) Rasterize ELC map

```{r}
ELC.sf <- sf::st_read("data/ELC_and_Land_Use_2022_-4248462527224701921/Ecological_Land_Classification_ELC_and_Land_Use_2022.shp") 
ELC.sf <- st_make_valid(ELC.sf )
Lower_WaterShed <- st_read(here::here("data/Lower_WaterShed/Lower_WaterShed.shp"))
```

```{r}
template.10 <- rast(extent=ext(vect(st_crop(ELC.sf, Lower_WaterShed))), 
                    resolution=10, crs=crs(vect(ELC.sf)))
LWS.r10 <- terra::rasterize(ELC.sf, template.10, "TYPE")
writeRaster(LWS.r10 , here::here("output/LWS.r10.tif"), overwrite=TRUE)
```


## b) Run `eco.landscape`

```{r eval=FALSE}
infile <- here::here("output/LWS.r10.tif")
output_folder <- here::here("output/LWS.r10")

Scales = c(500, 1000, 1500, 2000, 2500, 3000)
for(s in 1:length(Scales))
eco.landscape(
input_raster=infile,
scales=Scales[s],
classes=c(3:10),
output_folder=output_folder,
properties_file=file.path(output_folder, "params.properties"))
```

Notes:

- I had to run 'eco.landscape' separately for each scale because it produced only one tif file per #classes.
- The output .tif files had correct coordinates but incorrect crs information! A French crs was assigned. I had to import the files, change their crs information, and save them under a different name.

## c) Map results

Note: this is where I had to import the .tif files, change the CRS information, export, and re-import.

```{r}
Lower_WaterShed <- st_read(here::here("data/Lower_WaterShed/Lower_WaterShed.shp"))
output_folder <- here::here("output/LWS.r10")

mapfiles <- list.files(output_folder, pattern=".tif")
mapfiles <- mapfiles[-grep(pattern = "rupture", mapfiles)] 
outfiles <- gsub("ecopaysages_", "", mapfiles)
Ecolandscapes <- list()
for(i in 1:length(mapfiles))
{
  tmp <- terra::rast(file.path(output_folder, mapfiles[i]))
  terra::crs(tmp) <- terra::crs(vect(Lower_WaterShed))
  writeRaster(tmp, file.path(output_folder, outfiles[i]), overwrite=TRUE)
  Ecolandscapes[[i]] <- terra::rast(file.path(output_folder, outfiles[i]))
}

Ecolandscapes.LWS <- as.factor(terra::rast(Ecolandscapes)) # Stack the maps
```

```{r}
Ecolandscapes.LWS <- mask(Ecolandscapes.LWS, vect(Lower_WaterShed)), touches=FALSE)
#s = order(sapply(Ecolandscapes.m, function(r) nrow(levels(r)[[1]])))
#Ecolandscapes.m <- Ecolandscapes.m[[s]]

tmp <- t(simplify2array(strsplit(names(Ecolandscapes.LWS), split="_")))

dd <- data.frame(nClasses = as.numeric(gsub("\\D", "", tmp[,3])),
                 Scale = as.numeric(gsub("\\D", "", tmp[,4])))
dd$Label <- paste0(dd$Scale, " m, ", dd$nClasses, " groups")

s = with(dd, order(nClasses, Scale))
Ecolandscapes.LWS <- Ecolandscapes.LWS[[s]]
```

```{r}
tmap_mode("plot")
```

```{r fig.height=8, fig.width=7, message=FALSE, warning=FALSE}
myPalette <- c(palette=brewer.pal(n = 10, name = "Paired"))
#myPalette <- paste0(myPalette, "80")
#myPalette <- c("#00000000", myPalette)
tm_shape(Ecolandscapes.LWS) + 
  tm_raster(style="cat", palette=myPalette, legend.show=FALSE) +
  tm_facets(nrow=length(unique(dd$nClasses)), ncol=length(unique(dd$Scale))) +
  tm_layout(panel.labels = c(dd$Label[s]))
```

## d) Create interactive map

Create multipolygons from raster map

```{r}
m1 <- grep("1000m", names(Ecolandscapes.LWS))
m2 <- grep("6classes", names(Ecolandscapes.LWS))

LWS.Groups.1000m.6 <- st_as_sf(as.polygons(Ecolandscapes.LWS[[intersect(m1, m2)]]))
st_write(LWS.Groups.1000m.6, here::here("output/LWS.Groups.1000m.6.shp"), append=FALSE)
```

```{r}
tmap_mode("view")
```

```{r fig.height=5, fig.width=7}
tmap_options(basemaps=c(Imagery = "Esri.WorldImagery", Canvas = "Esri.WorldGrayCanvas"))
tm_shape(LWS.Groups.1000m.6) + 
  tm_sf("ecopaysages_LWS.r10_6classes_1000m", style="cat", alpha=0.5,
        legend.show=FALSE)
```

# 2. Repeat for upper watershed

## a) Rasterize ELC map

```{r}
ELC.sf <- sf::st_read("data/ELC_and_Land_Use_2022_-4248462527224701921/Ecological_Land_Classification_ELC_and_Land_Use_2022.shp") 
ELC.sf <- st_make_valid(ELC.sf )
Lower_WaterShed <- st_read(here::here("data/Lower_WaterShed/Lower_WaterShed.shp"))
CVC_Boundary <- st_read(here::here("data/CVC_boundaryt/CVC_boundaryt.shp"))
CVC_Boundary <- st_transform(CVC_Boundary, st_crs(Lower_WaterShed))
Upper_WaterShed <- st_difference(CVC_Boundary, st_buffer(Lower_WaterShed, 5))
```

```{r}
template.UWS.10 <- rast(extent=ext(vect(st_crop(ELC.sf, Upper_WaterShed))), 
                    resolution=10, crs=crs(vect(ELC.sf)))
UWS.r10 <- terra::rasterize(ELC.sf, template.UWS.10, "TYPE")
writeRaster(UWS.r10 , here::here("output/UWS.r10.tif"), overwrite=TRUE)
```


## b) Run `eco.landscape`

```{r eval=FALSE}
infile <- here::here("output/UWS.r10.tif")
output_folder <- here::here("output/UWS.r10")

Scales = c(500, 1000, 1500, 2000, 2500, 3000)
for(s in 1:length(Scales))
eco.landscape(
input_raster=infile,
scales=Scales[s],
classes=c(3:10),
output_folder=output_folder,
properties_file=file.path(output_folder, "params.properties"))
```


## c) Map results

```{r}
output_folder <- here::here("output/UWS.r10")

#file.remove(file.path(output_folder, outfiles))
mapfiles <- list.files(output_folder, pattern=".tif")
mapfiles <- mapfiles[-grep(pattern = "rupture", mapfiles)] 
outfiles <- gsub("ecopaysages_", "", mapfiles)
Ecolandscapes <- list()
for(i in 1:length(mapfiles))
{
  tmp <- terra::rast(file.path(output_folder, mapfiles[i]))
  terra::crs(tmp) <- terra::crs(vect(Upper_WaterShed))
  writeRaster(tmp, file.path(output_folder, outfiles[i]), overwrite=TRUE)
  Ecolandscapes[[i]] <- terra::rast(file.path(output_folder, outfiles[i]))
}

Ecolandscapes.UWS <- as.factor(terra::rast(Ecolandscapes)) # Stack the maps
```

```{r}
Ecolandscapes.UWS <- mask(Ecolandscapes.UWS, vect(Upper_WaterShed), touches=FALSE)
#s = order(sapply(Ecolandscapes.m, function(r) nrow(levels(r)[[1]])))
#Ecolandscapes.m <- Ecolandscapes.m[[s]]

tmp <- t(simplify2array(strsplit(names(Ecolandscapes.UWS), split="_")))

dd <- data.frame(nClasses = as.numeric(gsub("\\D", "", tmp[,3])),
                 Scale = as.numeric(gsub("\\D", "", tmp[,4])))
dd$Label <- paste0(dd$Scale, " m, ", dd$nClasses, " groups")

s = with(dd, order(nClasses, Scale))
Ecolandscapes.UWS <- Ecolandscapes.UWS[[s]]
```


```{r}
tmap_mode("plot")
```

```{r fig.height=8, fig.width=7, message=FALSE, warning=FALSE}
myPalette <- c(palette=brewer.pal(n = 10, name = "Paired"))
#myPalette <- paste0(myPalette, "80")
#myPalette <- c("#00000000", myPalette)
tm_shape(Ecolandscapes.UWS) + 
  tm_raster(style="cat", palette=myPalette, legend.show=FALSE) +
  tm_facets(nrow=length(unique(dd$nClasses)), ncol=length(unique(dd$Scale))) +
  tm_layout(panel.labels = c(dd$Label[s]))
```

Create multipolygons from raster map

```{r}
m1 <- grep("1500m", names(Ecolandscapes.UWS))
m2 <- grep("4classes", names(Ecolandscapes.UWS))

UWS.Groups.1500m.4 <- st_as_sf(as.polygons(Ecolandscapes.UWS[[intersect(m1, m2)]]))
st_write(UWS.Groups.1500m.4, here::here("output/UWS.Groups.1500m.4.shp"), append=FALSE)
```

# 3. Interpret delineated ecolandscapes

## a) Explore groups interactively

Import polygon maps of groups and renumber to avoid duplicate group numbers.

```{r}
tmap_mode("view")

UWS.Groups.1500m.4 <- st_read(here::here("output/UWS.Groups.1500m.4.shp")) 
UWS.Groups.1500m.4 <- UWS.Groups.1500m.4 %>% 
  rename(Group = names(UWS.Groups.1500m.4)[1]) %>%
  mutate(Group = as.numeric(Group))

LWS.Groups.1000m.6 <- st_read(here::here("output/LWS.Groups.1000m.6.shp"))
LWS.Groups.1000m.6 <- LWS.Groups.1000m.6 %>% 
  rename(Group = names(LWS.Groups.1000m.6)[1]) %>%
  mutate(Group = as.numeric(Group) + nrow(UWS.Groups.1500m.4))
```

Plot both maps (upper and lower watershed) together over satellite imagery to interactively interpret groups.

```{r fig.height=5, fig.width=7}
myPalette <- c(palette=brewer.pal(n = 10, name = "Paired"))
tmap_options(basemaps=c(Imagery = "Esri.WorldImagery", Canvas = "Esri.WorldGrayCanvas"))
tm_shape(UWS.Groups.1500m.4) + 
  tm_sf("Group", style="cat", alpha=0.5,
        legend.show=FALSE, palette=myPalette[1:4]) +
tm_shape(LWS.Groups.1000m.6) + 
  tm_sf("Group", style="cat", alpha=0.5,
        legend.show=FALSE, palette=myPalette[-c(1:4)])
```

## b) Select and label groups

Add attributes (variables): `Color`, `Keep`, `Label`

```{r}
Ecolandscapes_CVC <-  rbind(st_buffer(UWS.Groups.1500m.4, -10), 
                            st_buffer(LWS.Groups.1000m.6, -10)) %>%
  dplyr::mutate(Keep="N", Label=NA, Color=1:max(Group)) %>%
  dplyr::select(Color, Group, Keep, Label)
Ecolandscapes_CVC_2 <- Ecolandscapes_CVC
```

```{r}
tmap_mode("plot")
```

```{r message=FALSE, warning=FALSE}
tm_shape(Ecolandscapes_CVC) + 
  tm_sf("Group", style="cat", alpha=0.5, palette=myPalette) +
  tm_layout(legend.outside=TRUE)
```

```{r}
st_drop_geometry(Ecolandscapes_CVC_2)
```

For each group g:

- Decide whether it should be kept. If so, change the g^th value of `Keep` from "N" to "Y". 
- Change the g^th value of `Label` from "NA" to a meaningful label. 
- If two groups should be merged, assign them the same `Label` and `Color`.
- Optional: shuffle the Color numbers to reassign the ten colors to groups. 


```{r}
Keep <- c("N", "N", "N", "N", "N", "N", "N", "N", "N", "N")
Label <- c("NA", "NA", "NA", "NA", "NA", "NA", "NA", "NA", "NA", "NA")
Color <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```

```{r}
Keep <- c("Y", "Y", "Y", "Y", "Y", "Y", "Y", "Y", "Y", "Y")
Label <- c("Digs", "Rural", "More natural", "Rural settlement", 
            "Lakeshore", "Residential", "Commercial/Industrial", 
           "Rural", "Lakeshore", "River corridor")
Color <- c(1, 3, 4, 8, NA, 6, 10, 3, 9, 2)
#Color <- c(1, 7, 4, 6, NA, 8, 10, 7, 2, 3)
Color <- c(7, 3, 4, 8, NA, 5, 6, 3, 1, 2)
Color <- c(7, 3, 4, 8, 2, 5, 6, 3, 2, 10)
```

Check your edits:

```{r}
Ecolandscapes_CVC_2$Keep <- Keep
Ecolandscapes_CVC_2$Label <- Label
Ecolandscapes_CVC_2$Color <- Color
Ecolandscapes_CVC_2$colhex <- myPalette[Color]
st_drop_geometry(Ecolandscapes_CVC_2)
```

If you are satisfied with your edits, replot the map.

```{r}
Order <- c(3,2,4,1,7,6,10,9)
tmp <- Ecolandscapes_CVC_2 %>% dplyr::filter(Keep=="Y") %>%
           mutate(Label=factor(Label,levels=Label[Order]))

tm_shape(tmp) + 
  tm_fill(col="Label", alpha=0.8, palette=tmp$colhex[Order]) +
  tm_layout(main.title = "Ecolandscapes CVC",
          legend.position = c("right", "top"))
```

