---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


```{r}
# required packages
#install.packages("devtools")
#install.packages("Rtools") # Windows only?
# clean up before reloading a new version
#detach("package:chloe")
#.rs.restartR()

#remotes::install_github("pauI/chloeR") # install package
#library("chloe") # load package
get.java() # a Java Virtual Machine is needed, to retrieve the default path

#set.java("path/to/java/java.exe") # to declare a specific JVM
```

```{r}
library(chloe)
library(dplyr)
library(here)
library(terra)
library(tmap)
library(RColorBrewer)
library(sf)
```


# 1. Run example

## a) Run `eco.landscape`

Note: the `.tif` files in the output folder won't be overwritten. Need to either delete them first or create a new output subfolder.

```{r}
infile <- system.file("data", "sample.tif", package = "chloe")
output_folder <- here::here("output/output.test5")

eco.landscape(
input_raster=infile,
scales=c(1000),
classes=c(5, 6, 7),
output_folder=output_folder,
properties_file=file.path(output_folder, "params.properties"))
```

## b) Map results

```{r}
mapfiles <- list.files(output_folder, pattern=".tif")
mapfiles <- mapfiles[-grep(pattern = "rupture", mapfiles)] 
Ecolandscapes <- list()
for(i in 1:length(mapfiles))
{
  Ecolandscapes[[i]] <- terra::rast(file.path(output_folder, mapfiles[i]))
}
names(Ecolandscapes) <- substr(mapfiles, start=13, stop=33)
Ecolandscapes <- as.factor(terra::rast(Ecolandscapes)) # Stack the maps
```

```{r}
tmap_mode("plot")
tm_shape(Ecolandscapes) + tm_raster(legend.show=FALSE)
```

# 2. Apply to CVC ELC map

## a) Rasterize ELC map

```{r}
ELC.sf <- sf::st_read("data/ELC_and_Land_Use_2022_-4248462527224701921/Ecological_Land_Classification_ELC_and_Land_Use_2022.shp") 
ELC.sf <- st_make_valid(ELC.sf )
```

```{r}
template.30<- rast(extent=ext(vect(ELC.sf)), resolution=30, crs=crs(vect(ELC.sf)))
CVC.r30 <- terra::rasterize(ELC.sf, template.30, "TYPE")
template.10<- rast(extent=ext(vect(ELC.sf)), resolution=10, crs=crs(vect(ELC.sf)))
CVC.r10 <- terra::rasterize(ELC.sf, template.10, "TYPE")
writeRaster(CVC.r30 , here::here("output/CVC.r30.tif"), overwrite=TRUE)
writeRaster(CVC.r10 , here::here("output/CVC.r10.tif"), overwrite=TRUE)
```




## b) Run `eco.landscape`

```{r}
infile <- here::here("output/CVC.r10.tif")
output_folder <- here::here("output/CVC.r10")

eco.landscape(
input_raster=infile,
scales=c(1000),
classes=c(3, 5, 7, 10),
output_folder=output_folder,
properties_file=file.path(output_folder, "params.properties"))
```


## c) Map results

```{r}
mapfiles <- list.files(output_folder, pattern=".tif")
mapfiles <- mapfiles[-grep(pattern = "rupture", mapfiles)] 
outfiles <- gsub("ecopaysages_CVC", "CVC", mapfiles)
Ecolandscapes <- list()
for(i in 1:length(mapfiles))
{
  tmp <- terra::rast(file.path(output_folder, mapfiles[i]))
  terra::crs(tmp) <- terra::crs(CVC.r10)
  writeRaster(tmp, file.path(output_folder, outfiles[i]), overwrite=TRUE)
  Ecolandscapes[[i]] <- terra::rast(file.path(output_folder, outfiles[i]))
}
names(Ecolandscapes) <- substr(mapfiles, start=13, stop=33)
Ecolandscapes <- as.factor(terra::rast(Ecolandscapes)) # Stack the maps
```

```{r}
CVC_Boundary <- st_read(here::here("data/CVC_boundaryt/CVC_boundaryt.shp"))
Ecolandscapes.m <- mask(Ecolandscapes, vect(CVC_Boundary))
```


```{r}
tmap_mode("plot")
myPalette <- c(palette=brewer.pal(n = 10, name = "Paired"))
#myPalette <- paste0(myPalette, "80")
#myPalette <- c("#00000000", myPalette)
tm_shape(Ecolandscapes.m) + tm_raster(style="cat", palette=myPalette)
```


# 3. Apply to lower watershed ELC map

## a) Rasterize ELC map

```{r}
ELC.sf <- sf::st_read("data/ELC_and_Land_Use_2022_-4248462527224701921/Ecological_Land_Classification_ELC_and_Land_Use_2022.shp") 
ELC.sf <- st_make_valid(ELC.sf )
Lower_WaterShed <- st_read(here::here("data/Lower_WaterShed/Lower_WaterShed.shp"))
```

```{r}
template.10 <- rast(extent=ext(vect(st_crop(ELC.sf, Lower_WaterShed))), 
                    resolution=10, crs=crs(vect(ELC.sf)))
LWS.r10 <- terra::rasterize(ELC.sf, template.10, "TYPE")
writeRaster(LWS.r10 , here::here("output/LWS.r10.tif"), overwrite=TRUE)
```




## b) Run `eco.landscape`

```{r}
infile <- here::here("output/LWS.r10.tif")
output_folder <- here::here("output/LWS.r10")

Scales = c(500, 1000, 1500, 2000, 2500, 3000)
for(s in 1:length(Scales))
eco.landscape(
input_raster=infile,
scales=Scales[s],
classes=c(3:10),
output_folder=output_folder,
properties_file=file.path(output_folder, "params.properties"))
```

Notes:

- I had to run 'eco.landscape' separately for each scale because it produced only one tif file per #classes.
- The output .tif files had correct coordinates but incorrect crs information! A French crs was assigned. I had to import the files, change their crs information, and save them under a different name.

## c) Map results

```{r}
#Lower_WaterShed <- st_read(here::here("data/Lower_WaterShed/Lower_WaterShed.shp"))
#output_folder <- here::here("output/LWS.r10")

#file.remove(file.path(output_folder, outfiles))
mapfiles <- list.files(output_folder, pattern=".tif")
mapfiles <- mapfiles[-grep(pattern = "rupture", mapfiles)] 
outfiles <- gsub("ecopaysages_", "", mapfiles)
Ecolandscapes <- list()
for(i in 1:length(mapfiles))
{
  tmp <- terra::rast(file.path(output_folder, mapfiles[i]))
  terra::crs(tmp) <- terra::crs(vect(Lower_WaterShed))
  writeRaster(tmp, file.path(output_folder, outfiles[i]), overwrite=TRUE)
  Ecolandscapes[[i]] <- terra::rast(file.path(output_folder, outfiles[i]))
}

Ecolandscapes <- as.factor(terra::rast(Ecolandscapes)) # Stack the maps
```

```{r}
Ecolandscapes.m <- mask(Ecolandscapes, vect(Lower_WaterShed))
#s = order(sapply(Ecolandscapes.m, function(r) nrow(levels(r)[[1]])))
#Ecolandscapes.m <- Ecolandscapes.m[[s]]

tmp <- t(simplify2array(strsplit(names(Ecolandscapes.m), split="_")))

dd <- data.frame(nClasses = as.numeric(gsub("\\D", "", tmp[,3])),
                 Scale = as.numeric(gsub("\\D", "", tmp[,4])))
dd$Label <- paste0(dd$Scale, " m, ", dd$nClasses, " groups")

s = with(dd, order(nClasses, Scale))
Ecolandscapes.m <- Ecolandscapes.m[[s]]
```


```{r}
tmap_mode("plot")
myPalette <- c(palette=brewer.pal(n = 10, name = "Paired"))
#myPalette <- paste0(myPalette, "80")
#myPalette <- c("#00000000", myPalette)
tm_shape(Ecolandscapes.m) + 
  tm_raster(style="cat", palette=myPalette, legend.show=FALSE) +
  tm_facets(nrow=length(unique(dd$nClasses)), ncol=length(unique(dd$Scale))) +
  tm_layout(panel.labels = c(dd$Label[s]))
```

```{r}
m1 <- grep("1000m", names(Ecolandscapes.m))
m2 <- grep("6classes", names(Ecolandscapes.m))
tm_shape(Ecolandscapes.m[[intersect(m1, m2)]]) + 
  tm_raster(style="cat", palette=myPalette) +
  tm_layout(legend.outside=TRUE, title="Group") 
```

## d) Create interactive map

Create multipolygons from raster map

```{r}
m1 <- grep("1000m", names(Ecolandscapes.m))
m2 <- grep("6classes", names(Ecolandscapes.m))

Groups.1000m.6 <- st_as_sf(as.polygons(Ecolandscapes.m[[intersect(m1, m2)]]))
```

```{r}
tmap_mode("view")
tmap_options(basemaps=c(Imagery = "Esri.WorldImagery", Canvas = "Esri.WorldGrayCanvas"))
tm_shape(Groups.1000m.6) + 
  tm_sf("ecopaysages_LWS.r10_6classes_1000m", style="cat", alpha=0.5,
        legend.show=FALSE)
```


