---
title: "Remote Sensing Lab"
output: html_notebook
---

# 1. Introduction

## a) Goals:

- Understand how satellite imagery is converted into discrete or continuous raster data
- Link satellite imagery and derived maps of a known landscape (around home)
- Map characteristics of local landscape:
  - Quantitative variable: NDVI (measure of live biomass)
  - Categorical variable: LULC classification
- Compare satellite imagery from different sensors (different spatial resolution)

## b) Steps:

- Extract satellite imagery around a point of interest
- Display bands as true color and false color images
- Raster calculation:
  - NDVI
  - Urbanization index (optional)
- Supervised classification
  - Define legend: what will be mapped?
  - Create training set
  - Apply to entire local landscape
- Summarize landscape:
  - Continuous: mean
  - Discrete: percent of area
- Compare results among students
  - Rank local landscapes? (Positionality)
  - Correlation between NDVI and Urbanization?

## c) Sources: we will losely follow these tutorials:

- https://gisgeography.com/sentinel-2-bands-combinations/
- https://www.r-bloggers.com/2024/01/an-overview-of-the-rsi-r-package-for-retrieving-satellite-imagery-and-calculating-spectral-indices/
- https://urbanspatial.github.io/classifying_satellite_imagery_in_R/
- https://docs.digitalearthafrica.org/en/latest/sandbox/notebooks/Real_world_examples/Urban_index_comparison.html (drop this part?)

## d) Satellite data: 

Landsat 8 data from EarthExplorer (USGS)

- Note: may become unavailable after Aug 30, 2024?
- Needs a (free) account
- Instructions: https://gisgeography.com/usgs-earth-explorer-download-free-landsat-imagery/

Sentinel data

- Download instructions: https://gisgeography.com/how-to-download-sentinel-satellite-data/


## e) R packages

```{r}
#remotes::install_github("Permian-Global-Research/rsi")
```

```{r}
library(terra)
library(here)
library(tidyverse)
library(sf)
library(tmap)


# Only for me:
library(rsi)
library(pbmcapply)
library(parallel)
```

## 2. Assemble locations and imagery for class

### a) Assemble locations and create buffers

```{r}
files=list.files("~/Library/CloudStorage/OneDrive-UniversityofToronto/Teaching/BIO311_2024/Labs/Lab 1/StudentFiles/Locations/", full.names=TRUE)
```

```{r}
get_location <- function(f=files[1])
{
  tmp <- read.csv(f)
  return(tmp[1,])
}
```

```{r}
Locations <- lapply(files, function(f) get_location(f))

Locations.df <- data.frame(Label=unlist(lapply(Locations, function(ls) ls[1,1])),
                           Latitude=as.numeric(unlist(lapply(Locations, function(ls) ls[1,2]))),
                           Longitude=as.numeric(unlist(lapply(Locations, function(ls) ls[1,3]))))
View(Locations.df)
```

Fix issues:

```{r}
Locations.df$Label[nchar(Locations.df$Label) > 15] <- "NN"
Locations.df$Label <- gsub(" ","-", Locations.df$Label) 
```

Add my coordinates:

```{r}
Locations.df <- rbind(Locations.df, c(NA, 43.54910940, -79.6636611))
Locations.df$Label[nrow(Locations.df)] <- "UTM"
```

Convert to sf object, define and change projection

```{r}
Locations.latlon <- sf::st_as_sf(Locations.df, coords=c("Longitude", "Latitude"),
                                crs=4326)
Locations.sf <- sf::st_transform(Locations.latlon, crs=32617)
```

Create buffer of 1000 m radius around nearest intersection (first point in MyPoints)

```{r}
Buffers <- sf::st_buffer(Locations.sf, dist=1000)
```

Plot map to check location

```{r}
tmap_mode("view")
tm_shape(Buffers) + tm_sf(border.col="red", alpha=0)
```

### b) Get images: Landsat8

```{r}
landsat8_14Aug2024 <- lapply(1:nrow(Buffers), function(i) 
  try(get_landsat_imagery(Buffers[i,],
    start_date = "2024-08-14",
    end_date = "2024-08-15",
    output_filename = tempfile(fileext = ".tif")), silent=TRUE))
```

```{r}
a <- which(lapply(landsat8_14Aug2024,class) == "try-error")
for(i in 1:length(a))
{
  landsat8_14Aug2024[[a[i]]] <- try(get_landsat_imagery(Buffers[a[i],],
    start_date = "2024-08-20",
    end_date = "2024-08-21",
    output_filename = tempfile(fileext = ".tif")), silent=TRUE)
}
which(lapply(landsat8_14Aug2024,class) == "try-error")
```

```{r}
landsat8_14Aug2024 <- lapply(landsat8_14Aug2024, terra::rast)
```

### c) Get images: Sentinel2

```{r}
sentinel2_12Sept2024 <- lapply(1:nrow(Buffers), function(i) 
  try(get_sentinel2_imagery(Buffers[i,],
    start_date = "2024-09-12",
    end_date = "2024-09-13",
    output_filename = tempfile(fileext = ".tif")), silent=TRUE))
```

```{r}
a <- which(lapply(sentinel2_12Sept2024,class) == "try-error")
for(i in 1:length(a))
{
  sentinel2_12Sept2024[[a[i]]] <- try(get_sentinel2_imagery(Buffers[a[i],],
    start_date = "2024-09-03",
    end_date = "2024-09-04",
    output_filename = tempfile(fileext = ".tif")), silent=TRUE)
}
which(lapply(sentinel2_12Sept2024,class) == "try-error")
```

```{r}
sentinel2_12Sept2024 <- lapply(sentinel2_12Sept2024, terra::rast)
```

### c) Mask and export imagery

```{r}
sentinel2_12Sept2024 <- lapply(1:nrow(Buffers), function(i) 
  mask(sentinel2_12Sept2024[[i]], Buffers[i,]))
names(sentinel2_12Sept2024) <- Buffers$Label
landsat8_14Aug2024 <- lapply(1:nrow(Buffers), function(i) 
  mask(landsat8_14Aug2024[[i]], Buffers[i,]))
names(landsat8_14Aug2024) <- Buffers$Label
```


```{r}
par(mfrow = c(1, 2))
i=1
#i=i+1
terra::plotRGB(landsat8_14Aug2024[[i]], r = 4, g = 3, b = 2, stretch = "lin")
terra::plotRGB(sentinel2_12Sept2024[[i]], r = 4, g = 3, b = 2, stretch = "lin")
```

```{r}
lapply(1:nrow(Buffers), function(i) { 
 writeRaster(sentinel2_12Sept2024[[i]],  
   paste0("Images/sentinel2_12Sept2024_", Buffers$Label[i], ".tif"),
   overwrite=TRUE)
  }) 
```

```{r}
lapply(1:nrow(Buffers), function(i) { 
 writeRaster(landsat8_14Aug2024[[i]],  
   paste0("Images/landsat8_14Aug2024_", Buffers$Label[i], ".tif"),
   overwrite=TRUE)
  }) 
```


## 3. Display images

Extract imagery from lists

```{r}
MyPseudonym <- "UTM"
MySentinel2 <- readRDS("sentinel2_12Sept2024.rds")[[MyPseudonym]]
MyLandsat8 <- readRDS("landsat8_14Aug2024.rds")[[MyPseudonym]]
```


True-color display

```{r}
par(mfrow = c(1, 2))
i=1
#i=i+1
terra::plotRGB(landsat8_14Aug2024[[i]], r = 4, g = 3, b = 2, stretch = "lin")
terra::plotRGB(sentinel2_12Sept2024[[i]], r = 4, g = 3, b = 2, stretch = "lin")
```


Import formula to calculate NDVI

```{r}
ndvi = spectral_indices(download_indices = TRUE) %>%
  filter(short_name == "NDVI")
```

Apply to each sentinel2 scene

```{r}
sentinel2_12Sept2024_ndvi = calculate_indices(
    sentinel2_12Sept2024,
    ndvi,
    output_filename = tempfile(fileext = ".tif")
)
sentinel2_30June2024_ndvi = calculate_indices(
    sentinel2_30June2024,
    ndvi,
    output_filename = tempfile(fileext = ".tif")
)
sentinel2_30June2024_ndvi
```

```{r}
par(mfrow = c(1, 2))
sentinel2_12Sept2024_ndvi_rast = terra::rast(sentinel2_12Sept2024_ndvi)
sentinel2_30June2024_ndvi_rast = terra::rast(sentinel2_30June2024_ndvi)
terra::plot(sentinel2_12Sept2024_ndvi_rast, range = c(-1, 1))
terra::plot(sentinel2_30June2024_ndvi_rast, range = c(-1, 1))
```
Take difference in NDVI

```{r}
par(mfrow = c(1, 2))
dif = sentinel2_12Sept2024_ndvi_rast - sentinel2_30June2024_ndvi_rast
terra::plot(dif)
hist(dif, main = "", xlab = "NDVI")
```
Stack rasters, label them, mask with buffer (i.e., set outside area to NA)

```{r}
stack_rast = terra::rast(list("NDVI 05.30"=sentinel2_30June2024_ndvi_rast,
                              "NDVI 09.12"=sentinel2_12Sept2024_ndvi_rast))
stack_rast = mask(stack_rast, MyBuffer)
```


```{r}
terra::plot(stack_rast, range = c(-1, 1))
```

Calculate mean NDVI within buffer for each scene

```{r}
terra::global(stack_rast, mean, na.rm=TRUE)
```

K-means clustering to classify into vegetated / unvegetated / water:

https://rspatial.org/raster/rs/4-unsupclassification.html

```{r}
nr <- values(stack_rast$`NDVI 05.30`, mat=FALSE)
kmncluster <- kmeans(na.omit(nr), centers = 3, iter.max = 500, nstart = 5, algorithm="Lloyd")
nr[!is.na(nr)] <- kmncluster$cluster
knr <- setValues(stack_rast$`NDVI 05.30`, nr)
```

```{r}
mycolor <- c("lightgreen", "red3", "darkgreen")
par(mfrow = c(1,2))
plot(stack_rast$`NDVI 05.30`, col = rev(terrain.colors(10)), main = 'NDVI')
plot(knr, main = 'Unsupervised classification', col = mycolor )
#plot(sentinel2_12Sept2024, col = rev(terrain.colors(10)), main = 'Sentinel')
```


```{r}
landsat8_14Aug2024_ndvi = calculate_indices(
    landsat8_14Aug2024,
    ndvi,
    output_filename = tempfile(fileext = ".tif")
)
landsat8_14Aug2024_ndvi_rast = mask(terra::rast(landsat8_14Aug2024_ndvi), MyBuffer)
nr <- values(landsat8_14Aug2024_ndvi_rast, mat=FALSE)
kmncluster <- kmeans(na.omit(nr), centers = 3, iter.max = 500, nstart = 5, algorithm="Lloyd")
nr[!is.na(nr)] <- kmncluster$cluster
knr.Landsat8 <- setValues(landsat8_14Aug2024_ndvi_rast, nr)
```

```{r}
mycolor <- c("darkgreen", "lightgreen", "gray")
par(mfrow = c(1,2))
plot(landsat8_14Aug2024_ndvi_rast, col = rev(terrain.colors(10)), main = 'NDVI')
plot(knr.Landsat8, main = 'Unsupervised classification', col = mycolor )
#plot(sentinel2_12Sept2024, col = rev(terrain.colors(10)), main = 'Sentinel')
```
```{r}
terra::global(landsat8_14Aug2024_ndvi_rast, mean, na.rm=TRUE)
```


# 2. Explore Landsat Imagery

## a) Define area of interest (AOI)

Import points, convert to spatial feature with correct CRS, transform to UTM zone 17 CRS.

```{r}
MyPoints <- read.csv("MyPoints.csv")
MyPoints[1,2:3] <- c(43.66974504204339, -79.38948582208968)
MyPoints.latlon <- sf::st_as_sf(MyPoints, coords=c("Longitude", "Latitude"),
                                crs=4326)
MyPoints.utm <- sf::st_transform(MyPoints.latlon, crs=32617)
```

Create buffer of 1000 m radius around nearest intersection (first point in MyPoints)

```{r}
MyBuffer <- sf::st_buffer(MyPoints.utm[1,], dist=1000)
```

Plot map to check location

```{r}
tmap_mode("view")
tm_shape(MyBuffer) + tm_sf(border.col="red", alpha=0)
```


Import Landsat 8 scene

```{r}
datapath <- "data/Landsat8_14Aug2024"
filelist <- list.files(datapath, full.names = TRUE, pattern = "T1_B")
filelist <- filelist[which(is.na(str_extract(filelist, regex("B8.TIF"))))]
Image_bands <- rast(filelist)
names(Image_bands) <- as.character(sapply(names(Image_bands), function(x) substring(x, 42)))
Image_bands
```
Description of bands

```{r}
Bands <- data.frame(Band=paste0("B", 1:11), 
                    Description=c("Coastal/aerosol", 'Blue', 'Green',
                       'Red', 'Near Infrared (NIR)','Shortwave Infrared (SWIR) 1',
                       'Shortwave Infrared (SWIR) 2', 'Panchromatic', 'Cirrus',
                       'Thermal Infrared (TIRS) 1', 'Thermal Infrared (TIRS) 2' ))
Bands <- left_join(data.frame(Band=names(Image_bands)), Bands)
Bands
```

# 3. Raster Calculations

# 4. Supervised Classification

See also this resource: https://gisgeography.com/image-classification-techniques-remote-sensing/
Similar techniques are used in biological/medical imaging.



```{r}
library(mapedit)
library(mapview)
sentinel2_12Sept2024_brick <- raster::brick(sentinel2_12Sept2024)

```

Interactively select at least 10 points with vegetation

```{r}
#Vegetation.points <- viewRGB(sentinel2_12Sept2024_brick, r = 4, g = 3, b = 2) %>% 
Vegetation.points <- viewRGB(sentinel2_12Sept2024_brick, r = 8, g = 4, b = 3) %>% 
  editMap()
Vegetation <- Vegetation.points$finished$geometry %>% st_sf() %>% mutate(class = "Vegetation", id = 1)
```

Repeat for roads

```{r}
Road.points <- viewRGB(sentinel2_12Sept2024_brick, r = 8, g = 4, b = 3) %>% 
#Road.points <- viewRGB(sentinel2_12Sept2024_brick, r = 12, g = 8, b = 4) %>% 
  editMap()
if(length(Road.points$drawn) > 0)
{
  Road <- Road.points$finished$geometry %>% st_sf() %>% mutate(class = "Road", id = 2)
}else{
  Road <- Vegetation %>% filter(is.na(id))
}
```

Repeat for buildings

```{r}
Building.points <- viewRGB(sentinel2_12Sept2024_brick, r = 8, g = 4, b = 3) %>% 
#Building.points <- viewRGB(sentinel2_12Sept2024_brick, r = 11, g = 8, b = 2) %>% 
  editMap()
if(length(Building.points$drawn) > 0)
{
  Building <- Building.points$finished$geometry %>% st_sf() %>% mutate(class = "Building", id = 3)
}else{
  Building <- Vegetation %>% filter(is.na(id))
}
```


Repeat for water (if present)

```{r}
Water.points <- viewRGB(sentinel2_12Sept2024_brick, r = 8, g = 4, b = 3) %>% 
#Water.points <- viewRGB(sentinel2_12Sept2024_brick, r = 12, g = 8, b = 4) %>% 
  editMap()
if(length(Water.points$drawn) > 0)
{
  Water <- Water.points$finished$geometry %>% st_sf() %>% mutate(class = "Water", id = 4)
}else{
  Water <- Vegetation %>% filter(is.na(id))
}
```

Repeat for clouds (if present)

```{r}
Cloud.points <- viewRGB(sentinel2_12Sept2024_brick, r = 8, g = 4, b = 2) %>% 
  editMap()
if(length(Cloud.points$drawn) > 0)
{
  Cloud <- Cloud.points$finished$geometry %>% st_sf() %>% mutate(class = "Cloud", id = 5)
}else{
  Cloud <- Vegetation %>% filter(is.na(id))
}
```

```{r}
training_points <- rbind(Vegetation, Road, Building, Water, Cloud)
```


```{r}
training_points <- as(training_points, 'Spatial')

df <- raster::extract(sentinel2_12Sept2024_brick, training_points) %>%
  round()
```

Check for points with NA values in bands, remove them from training set

```{r}
a <- which(is.na(apply(df,1,sum)))
if(length(a) > 0)
{
  df <- df[!is.na(apply(df,1,sum)),]
  training_points <-  training_points[-a,]
}
```

```{r}
profiles <- df %>% 
  as.data.frame() %>% 
  cbind(., training_points$id) %>% 
  rename(id = "training_points$id") %>% 
  na.omit() %>% 
  group_by(id) %>% 
  summarise(A = mean(A),
            B = mean(B),
            G = mean(G),
            R = mean(R),
            RE1 = mean(RE1),
            RE2 = mean(RE2),
            RE3 = mean(RE3),
            N = mean(N),
            N2 = mean(N2),
            WV = mean(WV),
            S1 = mean(S1),
            S2 = mean(S2)) %>% 
  mutate(id = case_when(id == 1 ~ "Vegetation",
                        id == 2 ~ "Road",
                        id == 3 ~ "Building",
                        id == 4 ~ "Water",
                        id == 5 ~ "Cloud")) %>% 
  as.data.frame()

head(profiles)
```


```{r}
profiles %>% 
  select(-id) %>% 
  gather() %>% 
  mutate(class = rep(profiles$id, 12)) %>% 
  mutate(class = factor(class, levels=profiles$id)) %>%
  ggplot(., aes(x=value, group=as.factor(class), fill=as.factor(class))) + 
  geom_density(alpha = 0.75) + 
  geom_vline(data = . %>% group_by(class) %>% summarise(grp.mean = mean(value)),
             aes(xintercept=grp.mean, color = class), linetype="dashed", size=1) +
  scale_fill_manual(values=c('darkgreen', 'lightgreen', 'gray', 'red3','yellow', 'blue', 'white'), name = "class") +
  scale_color_manual(values=c('darkgreen', 'lightgreen', 'gray', 'red3','yellow', 'blue', 'white')) +
  theme(panel.background = element_blank(),
        panel.grid.major = element_line(color = "gray", size = 0.5),
        panel.grid.minor = element_line(color = "gray", size = 0.5),
        axis.ticks = element_blank()) +
  labs(x = "Reflectance Value",
       y = "Density",
       title = "Density histograms of spectral profiles",
       subtitle = "Vertical lines represent mean group reflectance values")
```

```{r}
library(rpart)
library(rpart.plot)
df <- data.frame(training_points.class=training_points$class, df)

model.class <- rpart(as.factor(training_points.class)~., data = df, method = 'class')

#cart <- rpart(as.factor(training_points.class)~., data=df, method = 'class', minsplit = 5)
rpart.plot(model.class, box.palette = 0, main = "Classification Tree")
```

```{r}
pr <- predict(sentinel2_12Sept2024_brick, model.class, type ='class', progress = 'text') %>% 
  raster::ratify()

levels(pr) <- levels(pr)[[1]] %>%
  mutate(legend = c("Road","Building","Vegetation"))
```

```{r}
rasterVis::levelplot(pr, col.regions = c('red2', 'orange', 'darkgreen'))
```



# 5. Compare Results among students

## a) Rank local landscapes

## b) Plot results in space

## c) Correlation between NDVI and Urbanization


