---
title: "Remote Sensing Lab"
output: html_notebook
---

# 1. Introduction

## a) Goals:

- Understand how satellite imagery is converted into discrete or continuous raster data
- Link satellite imagery and derived maps to known landscape (around home)
- Summarize local landscape

## b) Steps:

- Extract satellite imagery around a point of interest
  - Landsat 8 (30 m)
  - Sentinel 2 (10 m): 
    - Bands and combinations: https://gisgeography.com/sentinel-2-bands-combinations/. 
    - Analysis with R: https://bookdown.org/igisc/EnvDataSci/imagery.html
- Display bands as true color and false color images
- Raster calculation:
  - NDVI
- Unsupervised classification (optional)
- Supervised classification
  - Define legend: what will be mapped?
  - Create training set
  - Apply to entire local landscape
- Summarize landscape:
  - Continuous: mean
  - Discrete: percent of area
- Compare results among students
  - Rank local landscapes? (Positionality)
  - Correlation between NDVI and Urbanization?

## c) Sources: we will losely follow these tutorials:

- https://urbanspatial.github.io/classifying_satellite_imagery_in_R/
- https://docs.digitalearthafrica.org/en/latest/sandbox/notebooks/Real_world_examples/Urban_index_comparison.html

## d) Data: Landsat 8 data from EarthExplorer (USGS)

- Note: may become unavailable after Aug 30, 2024?
- Needs a (free) account
- Instructions: https://gisgeography.com/usgs-earth-explorer-download-free-landsat-imagery/

## e) R packages

```{r}
library(terra)
library(here)
library(tidyverse)
library(dplyr)
library(sf)
library(tmap)
library(stars)
```

# 2. Explore Landsat Imagery

## a) Import Landsat 8 scene

```{r}
datapath <- "data/Landsat8_14Aug2024"
filelist <- list.files(datapath, full.names = TRUE, pattern = "T1_B")
filelist <- filelist[which(is.na(str_extract(filelist, regex("B8.TIF"))))]
Image_bands <- rast(filelist)
names(Image_bands) <- as.character(sapply(names(Image_bands), function(x) substring(x, 42)))
Image_bands
```
## b) Description of bands

```{r}
Bands <- data.frame(Band=paste0("B", 1:11), 
                    Description=c("Coastal/aerosol", 'Blue', 'Green',
                       'Red', 'Near Infrared (NIR)','Shortwave Infrared (SWIR) 1',
                       'Shortwave Infrared (SWIR) 2', 'Panchromatic', 'Cirrus',
                       'Thermal Infrared (TIRS) 1', 'Thermal Infrared (TIRS) 2' ))
Bands <- left_join(data.frame(Band=names(Image_bands)), Bands)
Bands
```
Normalize values (range: 0 - 1)

```{r}
nx <- minmax(Image_bands, compute=TRUE)    
Image_rgb <- (Image_bands - nx[1,]) / (nx[2,] - nx[1,])
```

## c) Extract local landscape

Create spatial feature with point coordinates from Google Maps, Change transformation to match the coordinate system (crs) of the image.

```{r}
Point <-  data.frame(Latitude=43.66985782515024, Longitude=-79.38951424756152)
Point.sf <- st_as_sf(Point, coords=c("Longitude", "Latitude"), crs=st_crs(4326))
Point.sf <- st_transform(Point.sf, crs=st_crs(Image_bands))
```

Create a buffer around point, with a radius of 1000 m 

```{r}
Radius = 1000
Buffer <- st_buffer(Point.sf, Radius)
```

Limit data to buffer area (crop and mask)

```{r}
Landscape <- crop(Image_rgb, Buffer)
Landscape <-  mask(Landscape, Buffer)
```

```{r}
plotRGB(Landscape, r="B4", g="B3", b="B2", stretch="lin", smooth=FALSE)
```


# 3. Raster Calculations

## a) NDVI

```{r}
Landscape.NDVI <- (Landscape[["B5"]] - Landscape[["B4"]])/(Landscape[["B5"]] + Landscape[["B4"]]) 
plot(Landscape.NDVI)
```
```{r}
NDVI <- (Image_bands[["B5"]] - Image_bands[["B4"]])/(Image_bands[["B5"]] + Image_bands[["B4"]]) 
range(values(NDVI), na.rm = T)
```

```{r}
plot(NDVI)
```




```{r}
tm_shape(NDVI) + tm_raster(palette = "Greens")
```

```{r}
Landscape.NDVI <- mask(crop(NDVI, Buffer), Buffer)

tm_basemap(c(Imagery = "Esri.WorldImagery")) + 
tm_shape(Landscape.NDVI, raster.downsample=FALSE) + tm_raster(palette = "Greens", alpha=0.4)
```






# 4. Supervised Classification

See also this resource: https://gisgeography.com/image-classification-techniques-remote-sensing/
Similar techniques are used in biological/medical imaging.

# 5. Compare Results among students

## a) Rank local landscapes

## b) Plot results in space

## c) Correlation between NDVI and Urbanization



